Wait-free Queues with Polygarithmic Step Complexity

### 2022-07-08 article/psuedocode.tex
apply jul 7 feedbacks

\Function{boolean}{Refresh()}{}
\State \tt{h= \head} \label{readHead}
%\If{\tt{n.blocks[h]!=null?}} \tt{h+=1}\EndIf
\State \tt{<new, np\sub{left}, np\sub{right}>= \Call{CreateBlock}{s}} \cmt{\tt{np\sub{left}, np\sub{right}} are the values read from the children's \tt{num\sub{propagated}} field.}
\If{\tt{new.num==0}} \Return{\tt{true}} \label{addOP} \cmt{The block contains nothing.}
\ElsIf{\tt{blocks.tryAppend(new, h)}} \label{cas}
\ForEach{\tt{dir} {\keywordfont{in}} \tt{\{left, right\}}} \label{okcas}
\State \tt{\Call{CAS}{dir.super[np\sub{dir}], null, h}} \cmt{Write would work too.} \label{setSuper}
\State \tt{\Call{CAS}{dir.num\sub{propagated}, np\sub{dir}, np\sub{dir}+1}} \label{incNP}
\EndFor
\State \tt{\Call{CAS}{\head, h, h+1}} \label{incrementHead1}
\State \Return{\tt{true}}
\Else
\State \tt{\Call{CAS}{\head, h, h+1}} \cmt{Even if another process wins, help to increase the \tt{\head}. The winner might have fallen sleep before increasing \tt{\head}.} \label{incrementHead2}
\State \Return{ \tt{false}}
\EndIf
\EndFunction{Refresh}

@@

\begin{definition}[Subblock]\label{def::subblock}
Block \texttt{b} is a \emph{direct subblock} of \texttt{n.blocks[i]} if it is $\in$ \texttt{n.left.blocks[n.blocks[i-1].end\textsubscript{left}+1..n.blocks[i].end\textsubscript{left}]} $\cup$ \texttt{n.right.blocks[n.blocks[i-1].end\textsubscript{right}+1..n.blocks[i].end\textsubscript{right}]} (See line \ref{endDefLine} for the defined range). Block \texttt{b} is a subblock of a \texttt{n.blocks[i]} if it is a direct subblock of it or subblock of a direct subblock of it. 
\end{definition}

\begin{definition}[Superblock]
  Block b is direct superblock of block c if c is a direct subblock of b.   Block b is superblock of block c if c is a subblock of b.
\end{definition}

\begin{definition}[Operations of a block]\label{def::ops}
A block lb in a leaf represents one operation which if it is \tt{enqueue(x)} then \tt{lb.element=x}, otherwise \tt{element=null}. The set of operations of block $b$ are the operations in the subblocks of $b$. We show the set of operations of block b by \nf{ops(b)}.
\end{definition}

For simplicity we say block \nf{b} is propagated to node \nf{n} or to a set of blocks \nf{S} if \nf{b} is in \nf{n.blocks} or \nf{S} or is a subblock of a block in \nf{n.blocks} or \nf{S}. We also say $b$ contains $op$ if $op\in ops(b)$.

@@

\begin{definition}
 A block \nf{b} in \nf{n.blocks} is \emph{established} at time $t$ if the last value written into \nf{n.head} before $t$ is greater than the index of \nf{b} in \nf{n.blocks} at time $t$. \emph{EST\sub{n, t}} is the set of established blocks at time $t$ of node $n$.
\end{definition}

@@

\begin{lemma}[headProgress] \label{lem::headProgress}
\nf{n.head} is non-decreasing over time and \nf{n.blocks[i].end\sub{left}} $\geq$ \nf{n.blocks[i-1].end\sub{left}},\nf{n.blocks[i].end\sub{right}} $\geq$ \nf{n.blocks[i-1].end\sub{right}}.
\end{lemma}
\begin{proof}
The first claim follows trivially from the pseudocode since \nf{n.head} is only incremented. Also  when \nf{n.blocks[i]} is created its \nf{end\sub{left}, end\sub{right}} are greater than or equal to the values in \nf{n.blocks[i-1]}. Since \nf{blocks[i-1].end\sub{dir}}$<$\nf{dir.\head = blocks[i].end\sub{dir}} (Lines~\ref{lastLine, prevLine}).
\end{proof}

>>>>

\Function{boolean}{Refresh()}{}
\State \tt{h= \head} \label{readHead}
%\If{\tt{n.blocks[h]!=null?}} \tt{h+=1}\EndIf
\State \tt{<new, np\sub{left}, np\sub{right}>= \Call{CreateBlock}{h}} \label{invokeCreateBlock} \cmt{\tt{np\sub{left}, np\sub{right}} are the values read from the children's \tt{num\sub{propagated}} field.}
\If{\tt{new.num==0}} \Return{\tt{true}} \label{addOP} \cmt{The block contains nothing.}
\ElsIf{\tt{blocks.tryAppend(new, h)}} \label{cas}
\ForEach{\tt{dir} {\keywordfont{in}} \tt{\{left, right\}}} \label{okcas}
\State \tt{\Call{CAS}{dir.super[np\sub{dir}], null, h}} \cmt{Write would work too.} \label{setSuper}
\State \tt{\Call{CAS}{dir.num\sub{propagated}, np\sub{dir}, np\sub{dir}+1}} \label{incNP}
\EndFor
\State \tt{\Call{CAS}{\head, h, h+1}} \label{incrementHead1}
\State \Return{\tt{true}}
\Else
\State \tt{\Call{CAS}{\head, h, h+1}} \cmt{Even if another process wins, help to increase the \tt{\head}. The winner might have fallen sleep before increasing \tt{\head}.} \label{incrementHead2}
\State \Return{ \tt{false}}
\EndIf
\EndFunction{Refresh}

@@

\begin{lemma}[headProgress] \label{lem::headProgress}
\nf{n.head} is non-decreasing over time and \nf{n.blocks[i].end\sub{left}} $\geq$ \nf{n.blocks[i-1].end\sub{left}} and \nf{n.blocks[i].end\sub{right}} $\geq$ \nf{n.blocks[i-1].end\sub{right}}.
\end{lemma}
\begin{proof}
The first claim follows trivially from the pseudocode since \nf{n.head} is only incremented in the pseudocode in the lines \ref{incrementHead1} and \ref{incrementHead2} of the \nf{Refresh()} .

Consider the block \nf{b} written into \nf{n.blocks[i]} by \nf{TryAppend()} at Line \ref{cas}. It is created by the \nf{CreateBlock()} called at Line \ref{invokeCreateBlock}. Prior to this call to \nf{CreateBlock()}, \nf{n.head=i} at Line \ref{readHead}, so \nf{n.blocks[i-1]} is already a non-null value \nf{b}$^\prime$ by invariant \ref{lem::headPosition}. Thus the \nf{CreateBlock(i-1)} that creates  \nf{b}$^\prime$ terminates before \nf{CreateBlock(i)} that creates \nf{b} is invoked. The value written into \nf{b.end\sub{left}} at Line \ref{endDefLine} of \nf{CreateBlock()}  was read from \nf{n.left.head} at Line \ref{lastLine} of \nf{CreateBlock}. Similarly, the value in \nf{n.blocks[i-1].end\sub{left}} was read from \nf{n.left.head}. Since \nf{n.left.head} is non-decreasing \nf{b}$^\prime$\nf{.end\sub{left}} $\leq$ \nf{b.end\sub{left}}. The proof for \nf{end\sub{right}} is similar.
\end{proof}

\begin{definition}[Subblock]\label{def::subblock}
Block \texttt{b} is a \emph{direct subblock} of \texttt{n.blocks[i]} if it is in \texttt{n.left.blocks[n.blocks[i-1].end\textsubscript{left}+1..n.blocks[i].end\textsubscript{left}]} $\cup$ \texttt{n.right.blocks[n.blocks[i-1].end\textsubscript{right}+1..n.blocks[i].end\textsubscript{right}]} (The \nf{end\sub{left}, end\sub{right}} fields are written in \ref{endDefLine} of \nf{CreateBlock()}). Block \texttt{b} is a subblock of \texttt{n.blocks[i]} if \nf{b} is a direct subblock of \texttt{n.blocks[i]} or a subblock of a direct subblock of \texttt{n.blocks[i]}. 
\end{definition}

\begin{definition}[Superblock]
  Block \nf{b} is \t{direct superblock} of block \nf{c} if c is a direct subblock of b.   Block \nf{b} is \it{superblock} of block \nf{c} if \nf{c} is a subblock of \nf{b}.
\end{definition}

\begin{definition}[Operations of a block]\label{def::ops}
A leaf block \nf{b} in a leaf represents operation \nf{x} if \tt{b.element=x$\neq$null}. The set of operations of block \nf{b} are the operations in the subblocks of \nf{b}. We denote the set of operations of block \nf{b} by \nf{ops(b)}.

We say block \nf{b} is \it{propagated to node} \nf{n} if \nf{b} is in \nf{n.blocks} or is a subblock of a block in \nf{n.blocks}. We also say \nf{b} contains \nf{op} if \nf{op}$\in$\nf{ops(b)}.
\end{definition}

@@

\begin{definition}
 A block \nf{b} in \nf{n.blocks} is \emph{established} at time $t$ if the last value written into \nf{n.head} before $t$ is greater than the index of \nf{b} in \nf{n.blocks} at time $t$. \emph{EST\sub{n, t}} is the set of established blocks  of node \nf{n} at time $t$.
\end{definition}

@@
